---
title: "Breast Cancer Prediction"
author: "134780 Trevor Okinda"
date: "2024"
output:
  github_document: 
    toc: yes
    toc_depth: 4
    fig_width: 6
    fig_height: 4
    df_print: default
editor_options:
  chunk_output_type: console
---

# Student Details

|                                              |     |
|----------------------------------------------|-----|
| **Student ID Number**                        | 134780 |
| **Student Name**                             | Trevor Okinda |
| **BBIT 4.2 Group**                           | C |
| **Project Name**                             | Breast Cancer Prediction |

# Setup Chunk

**Note:** the following KnitR options have been set as the global defaults: <BR> `knitr::opts_chunk$set(echo = TRUE, warning = FALSE, eval = TRUE, collapse = FALSE, tidy = TRUE)`.

More KnitR options are documented here <https://bookdown.org/yihui/rmarkdown-cookbook/chunk-options.html> and here <https://yihui.org/knitr/options/>.

```{r setup, include=FALSE}
library(formatR)
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = FALSE
)
```

### Source: 

The dataset that was used can be downloaded here: *\<https://www.kaggle.com/datasets/merishnasuwal/breast-cancer-prediction-dataset\>*

### Reference:

*\<merishnasuwal. (n.d.). Breast Cancer Prediction Dataset. Kaggle. Retrieved April 17, 2025, from https://www.kaggle.com/datasets/merishnasuwal/breast-cancer-prediction-dataset\>\
Refer to the APA 7th edition manual for rules on how to cite datasets: <https://apastyle.apa.org/style-grammar-guidelines/references/examples/data-set-references>*

# Understanding the Dataset (Exploratory Data Analysis (EDA))
## Loading dataset
```{r loading dataset}
# Load dataset
breast_cancer_data <- read.csv("Breast_cancer_data.csv", colClasses = c(
  mean_radius = "numeric",
  mean_texture = "numeric",
  mean_perimeter = "numeric",
  mean_area = "numeric",
  mean_smoothness = "numeric",
  diagnosis = "factor"
))

# Display the structure of the dataset
str(breast_cancer_data)

# View the first few rows of the dataset
head(breast_cancer_data)

# View the dataset in a separate viewer window
View(breast_cancer_data)
```

## Measures of Frequency
```{r MOF}
# Frequency distribution for the diagnosis
diagnosis_freq <- table(breast_cancer_data$diagnosis)
print(diagnosis_freq)

# Proportion of each class
diagnosis_prop <- prop.table(diagnosis_freq)
print(diagnosis_prop)
```

## Measures of Central Tendency
```{r MOCT}
# Mean for numeric variables
mean_values <- sapply(breast_cancer_data[, 1:5], mean)
print(mean_values)

# Median for numeric variables
median_values <- sapply(breast_cancer_data[, 1:5], median)
print(median_values)

# Mode function
get_mode <- function(v) {
  uniqv <- unique(v)
  uniqv[which.max(tabulate(match(v, uniqv)))]
}
```

## Measures of Distribution
```{r MOD}
# Since mode is not typically used for numeric variables in this context, we'll skip it for now.
# Variance for numeric variables
variance_values <- sapply(breast_cancer_data[, 1:5], var)
print(variance_values)

# Standard deviation for numeric variables
sd_values <- sapply(breast_cancer_data[, 1:5], sd)
print(sd_values)

# Skewness and kurtosis require the 'moments' package
library(moments)

# Skewness for numeric variables
skewness_values <- sapply(breast_cancer_data[, 1:5], skewness)
print(skewness_values)

# Kurtosis for numeric variables
kurtosis_values <- sapply(breast_cancer_data[, 1:5], kurtosis)
print(kurtosis_values)
```

## Measures of Relationship
```{r MOR}
# Correlation matrix for numeric variables
correlation_matrix <- cor(breast_cancer_data[, 1:5])
print(correlation_matrix)

# Scatter plot matrix
pairs(breast_cancer_data[, 1:5], main = "Scatterplot Matrix")
```

## ANOVA
```{r ANOVA}
# ANOVA for mean_radius
anova_mean_radius <- summary(aov(mean_radius ~ diagnosis, data = breast_cancer_data))
print(anova_mean_radius)

# ANOVA for mean_texture
anova_mean_texture <- summary(aov(mean_texture ~ diagnosis, data = breast_cancer_data))
print(anova_mean_texture)

# ANOVA for mean_perimeter
anova_mean_perimeter <- summary(aov(mean_perimeter ~ diagnosis, data = breast_cancer_data))
print(anova_mean_perimeter)

# ANOVA for mean_area
anova_mean_area <- summary(aov(mean_area ~ diagnosis, data = breast_cancer_data))
print(anova_mean_area)

# ANOVA for mean_smoothness
anova_mean_smoothness <- summary(aov(mean_smoothness ~ diagnosis, data = breast_cancer_data))
print(anova_mean_smoothness)
```

## Plots
```{r Plots}
# Load ggplot2 for plotting
library(ggplot2)

# Create histograms for numeric variables
ggplot(breast_cancer_data, aes(x = mean_radius)) + 
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  ggtitle("Histogram of Mean Radius")

ggplot(breast_cancer_data, aes(x = mean_texture)) + 
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  ggtitle("Histogram of Mean Texture")

ggplot(breast_cancer_data, aes(x = mean_perimeter)) + 
  geom_histogram(binwidth = 5, fill = "blue", color = "black") +
  ggtitle("Histogram of Mean Perimeter")

ggplot(breast_cancer_data, aes(x = mean_area)) + 
  geom_histogram(binwidth = 50, fill = "blue", color = "black") +
  ggtitle("Histogram of Mean Area")

ggplot(breast_cancer_data, aes(x = mean_smoothness)) + 
  geom_histogram(binwidth = 0.01, fill = "blue", color = "black") +
  ggtitle("Histogram of Mean Smoothness")
# Create box plots for numeric variables by diagnosis
ggplot(breast_cancer_data, aes(x = diagnosis, y = mean_radius, fill = diagnosis)) + 
  geom_boxplot() +
  ggtitle("Box Plot of Mean Radius by Diagnosis")

ggplot(breast_cancer_data, aes(x = diagnosis, y = mean_texture, fill = diagnosis)) + 
  geom_boxplot() +
  ggtitle("Box Plot of Mean Texture by Diagnosis")

ggplot(breast_cancer_data, aes(x = diagnosis, y = mean_perimeter, fill = diagnosis)) + 
  geom_boxplot() +
  ggtitle("Box Plot of Mean Perimeter by Diagnosis")

ggplot(breast_cancer_data, aes(x = diagnosis, y = mean_area, fill = diagnosis)) + 
  geom_boxplot() +
  ggtitle("Box Plot of Mean Area by Diagnosis")

ggplot(breast_cancer_data, aes(x = diagnosis, y = mean_smoothness, fill = diagnosis)) + 
  geom_boxplot() +
  ggtitle("Box Plot of Mean Smoothness by Diagnosis")

# Bar plot for the diagnosis variable
ggplot(breast_cancer_data, aes(x = diagnosis, fill = diagnosis)) + 
  geom_bar() +
  ggtitle("Bar Plot of Diagnosis")

# Scatter plots colored by diagnosis
ggplot(breast_cancer_data, aes(x = mean_radius, y = mean_texture, color = diagnosis)) + 
  geom_point() +
  ggtitle("Scatter Plot of Mean Radius vs. Mean Texture")

ggplot(breast_cancer_data, aes(x = mean_radius, y = mean_perimeter, color = diagnosis)) + 
  geom_point() +
  ggtitle("Scatter Plot of Mean Radius vs. Mean Perimeter")

ggplot(breast_cancer_data, aes(x = mean_radius, y = mean_area, color = diagnosis)) + 
  geom_point() +
  ggtitle("Scatter Plot of Mean Radius vs. Mean Area")

ggplot(breast_cancer_data, aes(x = mean_radius, y = mean_smoothness, color = diagnosis)) + 
  geom_point() +
  ggtitle("Scatter Plot of Mean Radius vs. Mean Smoothness")

# Install and load GGally for pair plots
library(GGally)

# Pair plot for numeric variables colored by diagnosis
ggpairs(breast_cancer_data, columns = 1:5, ggplot2::aes(color = diagnosis))

```

# Preprocessing and Data Transformation
## Missing Values
```{r Missing Values}
# Check for missing values in the entire dataset
total_missing_values <- sum(is.na(breast_cancer_data))
print(paste("Total missing values:", total_missing_values))

# Install and load the VIM package
library(VIM)

# Plot missing data
aggr(breast_cancer_data, col = c('navyblue', 'red'), numbers = TRUE, sortVars = TRUE, labels = names(breast_cancer_data), cex.axis = .7, gap = 3, ylab = c("Missing data", "Pattern"))


```

# Training Model
## Data Splitting
```{r Data Splitting}
# Install and load the caTools package
library(caTools)

# Set seed for reproducibility
set.seed(123)

# Split the data into training (70%) and testing (30%) sets
split <- sample.split(breast_cancer_data$diagnosis, SplitRatio = 0.7)
training_set <- subset(breast_cancer_data, split == TRUE)
testing_set <- subset(breast_cancer_data, split == FALSE)

# Display the structure of the training and testing sets
str(training_set)
str(testing_set)
dim(training_set)
dim(testing_set)
```

## Bootstrapping
```{r Bootstrapping}
# Install and load the boot package
library(boot)

# Define a function to compute the statistic of interest (mean in this case)
bootstrap_function <- function(data, indices) {
  d <- data[indices, ]  # allows boot to select sample
  return(mean(d$mean_radius))  # example statistic
}

# Apply bootstrapping
set.seed(123)
bootstrap_results <- boot(data = breast_cancer_data, statistic = bootstrap_function, R = 1000)

# Display bootstrapping results
print(bootstrap_results)
plot(bootstrap_results)
```

## Cross-validation
```{r Cross-validation}
# Install and load the caret package
library(caret)

# Set up cross-validation
train_control <- trainControl(method = "cv", number = 10)

# Train a model using cross-validation (example with logistic regression)
set.seed(123)
model <- train(diagnosis ~ mean_radius + mean_texture + mean_perimeter + mean_area + mean_smoothness,
               data = training_set,
               method = "glm",
               family = binomial,
               trControl = train_control)

# Print the model summary
print(model)

# Make predictions on the testing set
predictions <- predict(model, newdata = testing_set)

# Evaluate the model's performance
confusion_matrix <- confusionMatrix(predictions, testing_set$diagnosis)
print(confusion_matrix)
```

# Model Training
```{r Model Training}
# Model Training

# Generalized Linear Model (GLM)
# Install and load the caret package
library(caret)

# Train a GLM model
set.seed(123)
glm_model <- train(diagnosis ~ mean_radius + mean_texture + mean_perimeter + mean_area + mean_smoothness,
                   data = training_set,
                   method = "glm",
                   family = binomial,
                   trControl = trainControl(method = "cv", number = 10))

# Print the GLM model summary
print(glm_model)

# Make predictions on the testing set
glm_predictions <- predict(glm_model, newdata = testing_set)

# Evaluate the GLM model's performance
glm_confusion_matrix <- confusionMatrix(glm_predictions, testing_set$diagnosis)
print(glm_confusion_matrix)

# Support Vector Machine (SVM)
# Install and load the e1071 package for SVM
library(e1071)

# Train a SVM model
set.seed(123)
svm_model <- train(diagnosis ~ mean_radius + mean_texture + mean_perimeter + mean_area + mean_smoothness,
                   data = training_set,
                   method = "svmLinear",
                   trControl = trainControl(method = "cv", number = 10))

# Print the SVM model summary
print(svm_model)

# Make predictions on the testing set
svm_predictions <- predict(svm_model, newdata = testing_set)

# Evaluate the SVM model's performance
svm_confusion_matrix <- confusionMatrix(svm_predictions, testing_set$diagnosis)
print(svm_confusion_matrix)

# Principal Component Analysis (PCA) with Logistic Regression
# Perform PCA
pca <- prcomp(training_set[, 1:5], center = TRUE, scale. = TRUE)

# Print the summary of PCA
summary(pca)

# Add the PCA results to the training and testing sets
training_set_pca <- as.data.frame(predict(pca, training_set[, 1:5]))
training_set_pca$diagnosis <- training_set$diagnosis

testing_set_pca <- as.data.frame(predict(pca, testing_set[, 1:5]))
testing_set_pca$diagnosis <- testing_set$diagnosis

# Train a logistic regression model on the PCA components
set.seed(123)
pca_glm_model <- train(diagnosis ~ PC1 + PC2 + PC3 + PC4 + PC5,
                       data = training_set_pca,
                       method = "glm",
                       family = binomial,
                       trControl = trainControl(method = "cv", number = 10))

# Print the PCA-GLM model summary
print(pca_glm_model)

# Make predictions on the testing set using the PCA components
pca_glm_predictions <- predict(pca_glm_model, newdata = testing_set_pca)

# Evaluate the PCA-GLM model's performance
pca_glm_confusion_matrix <- confusionMatrix(pca_glm_predictions, testing_set_pca$diagnosis)
print(pca_glm_confusion_matrix)
```

# Saving Model
```{r Saving Model}
# Load the saved SVM model
loaded_model_svm <- readRDS("./models/svm_model.rds")

# Prepare new data for prediction
new_data <- data.frame(
  mean_radius = 17.99,
  mean_texture = 10.38,
  mean_perimeter = 122.8,
  mean_area = 1001,
  mean_smoothness = 0.1184
)

# Use the loaded model to make predictions for new data
predictions_loaded_model <- predict(loaded_model_svm, newdata = new_data)

# Print predictions
print(predictions_loaded_model)

```


